# Appendix

## Table of operators

| Operator | Definition |
| :------: | :--------- |
|    ==    | Equal to                      |
|    !=    | Not equal to                  |
|    >     | Greater than                  |
|    <     | Less than                     |
|    >=    | Greater than or equal to      |
|    <=    | Less than or equal to         |
|  &#448;  | Or                            |
|    &     | And                           |
|    !     | Not                           |
|   %in%   | The subject appears in a list |
| is.na()  | The subject is NA             |




## Regular Expressions

Matching patterns with regular expressions is a large topic and one that could be a course of its own. Here we include a flavor of what can be achieved.


### Normal Pattern Matching 

Simple pattern matching can be achieved by using the `str_view` function and specifying the `string` and `pattern` arguments as shown below:



```{r, results=F}
x <- c("apple", "banana", "pear")
# str_view(x, "an")
```


The complexity of the match can be adjusted and wildcards can be used as well in the form of `.` as in 

```{r, results=F}
# str_view(x, ".a.")
```


An important thing to remember here is that you are looking for given pattern in a string or a vector of strings. Specifying the pattern to look for can be sometimes tricky and therefore is recommended to take a look at the accompanied information sheet for the `stringr` package located [here](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_strings.pdf). 

The most common functions for regular expressions as well as ways to structure your pattern can be found in the above link.

### Escaping characters

We saw previously that the wildcard is represented by `.`, what happens if your pattern needs to include the dot `.`?

For cases such as these there is a need to "escape" a character using the backlash `\`. So for example the dot (`.`) as stated previously would be normally used as `\.` in a regular expression. However, a problem arises since strings are used to represent regular expressions, they also use the backlash to represent an escaped character. The solution is to use the double backlash as in `\\.` to signify that we want to "escape" the dot (`.`) in a regular expression pattern.


```{r, results=F}
# To create the regular expression, we need \\
dot <- "\\."
# But the expression itself only contains one:
writeLines(dot)
# And this tells R to look for an explicit .
# str_view(c("abc", "a.c", "bef"), "a\\.c")
```


If there is a need to match the `\` character itself then you will need to use the double version `\\` for regular expressions and since this is a string you will need to add one more `\` followed by the actual character. So four backlash characters will need to be used!


```{r}
#to see this in a string
x <- "a\\b"
writeLines(x)
# to view the result in a RegEx
# str_view(x, "\\\\")
```


throughout this section the pattern for a RegEx will be presented as `\.` whereas the actual string as `\\.`. 



#### Exercises


1. Explain why each of these strings don’t match a \: "\", "\\", "\\\".

2. How would you match the sequence "'\?

3. What patterns will the regular expression \..\..\.. match? How would you represent it as a string?



### Anchors


It is sometimes useful to match a pattern starting from the beginning or the end of a string. In these cases an anchor is used to notify the engine that we are using a point of origin. 

More specifically, the start of the string is denoted by `^` and the end by `$`

```{r, results=F}
x <- c("apple", "banana", "pear")
# str_view(x, "^a")
# str_view(x, "a$")
```


You can also use both in one pattern and this is useful when the entire string is to be matched 

```{r, results=F}
# this will output all possible matches
x <- c("apple pie", "apple", "apple cake")
# str_view(x, "apple")
# notice the difference in the result here
# str_view(x, "^apple$")
```



#### Exercises

1. How would you match the literal string "$^$"?

2. Given the corpus of common words in stringr::words, create regular expressions that find all words that:

      +Start with “y”.
      +End with “x”
      +Are exactly three letters long. (Don’t cheat by using str_length()!)
      +Have seven letters or more.
Hint: Since this list is long, you might want to use the match argument to str_view() to show only the matching or non-matching words.

### Character Classes

Similar to the wildcard you saw previously, there are other reserved patters that serve a similar purpose for example:
 
 + `\d`: matches any digit.
 + `\s`: matches any whitespace (e.g. space, tab, newline).
 + `[abc]`: matches a, b, or c.
 + `[^abc]`: matches anything except a, b, or c.

A reminder here that if you want to use the above to pattern then you will need to escape them as we learned earlier. So the `\d` string would be used as `\\d` in a pattern.

In addition, there is an alternative to the backlash way of escaping a character that involves creating a class for a single character as in `[.]` for example. In many cases this is considered more intuitive that using the backlash.


A character class containing a single character is a nice alternative to backslash escapes when you want to include a single meta-character in a regex. Many people find this more readable.

```{r, results= F }
# Look for a literal character that normally has special meaning in a regex
# str_view(c("abc", "a.c", "a*c", "a c"), "a[.]c")
# str_view(c("abc", "a.c", "a*c", "a c"), ".[*]c")
# str_view(c("abc", "a.c", "a*c", "a c"), "a[ ]")
```


However, some characters will have a certain meaning even inside brackets and so the backslash for escaping them is still necessary, these are `] \ ^ -`.

Alternation is used to pick between patterns, for example `abc|d..f` will pick a pattern with either `c` or `d` in the definition. Note how the `|` is used, it will only pick between the immediate characters and not between the entire patterns on either side. If it gets to a point where it becomes confusing remember to use parenthesis to clear things up. 


```{r, results = F}
# str_view(c("grey", "gray"), "gr(e|a)y")
```



#### Exercises

1. Create regular expressions to find all words that:
    +Start with a vowel.
    +That only contain consonants. (Hint: thinking about matching “not”-vowels.)
    +End with` ed`, but not with `eed`.
    +End with `ing` or` ise`.
2. Empirically verify the rule “i before e except after c”.
3. Is “q” always followed by a “u”?
4. Write a regular expression that matches a word if it’s probably written in British English, not American English.
5. Create a regular expression that will match telephone numbers as commonly written in your country.

### Repetition

It is sometimes necessary for a certain pattern to appear multiple time within a string, in such cases these repetitions can be coded with regular expressions to automate the search process. 

for example:

+ `?`: 0 or 1
+ `+`: 1 or more
+ `*`: 0 or more

```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"
# str_view(x, "CC?")
# str_view(x, "CC+")
# str_view(x, 'C[LX]+')
```

Another key aspect of the above code is that the number or precedence here dictates that the character just before the operator will be affected. This means that parenthesis will need to be used as in `bana(na)+` to capture more than one character.

It is also possible to specify the number of repetitions explicitly by using:

+ `{n}`: exactly n
+ `{n,}`: n or more
+ `{,m}`: at most m
+ `{n,m}`: between n and m

```{r}
# str_view(x, "C{2}")
# str_view(x, "C{2,}")
# str_view(x, "C{2,3}")
```


To also note here that the system will match as many of the characters that it can find. To switch this behavior off and use what is called "lazy" matching (instead of "greedy" as specified earlier) the `?` operator can be used as follows:

```{r}
# str_view(x, 'C{2,3}?')
# str_view(x, 'C[LX]+?')
```

#### Exercises


1. Describe the equivalents of `?`, `+`,` *` in `{m,n}` form.

2. Describe in words what these regular expressions match: (read carefully to see if I’m using a regular expression or a string that defines a regular expression.)

      + `^.*$`
      + `"\\{.+\\}"`
      + `\d{4}-\d{2}-\d{2}`
      + `"\\\\{4}"`
3. Create regular expressions to find all words that:

      + Start with three consonants.
      + Have three or more vowels in a row.
      + Have two or more vowel-consonant pairs in a row.
      
4. Solve the beginner regexp crosswords at [https://regexcrossword.com/challenges/beginner].




