# Conditional statements

## if...else statements

Conditional statements can be used when you want a piece of code to be executed only if a particular condition is met. The most basic form of these are 'if...else' statements. As a simple example, let's say we wanted to check if a variable `x` is less than 10. We can write something like:

```{r, purl=purl_example_code}
x <- 9

if (x < 10) {
  print("x is less than 10")
}
```

We can also specify if we want something different to happen if the condition is not met:
```{r, purl=purl_example_code}
x <- 11

if (x < 10) {
  print("x is less than 10")
} else {
  print("x is 10 or greater")
}
```
Or if there are multiple conditions where we want different things to happen, we can add 'else if' commands:
```{r, purl=purl_example_code}
x <- 10

if (x < 10) {
  print("x is less than 10")
} else if (x == 10) {
  print("x is equal to 10")
} else if (x > 10) {
  print("x is greater than 10")
} else {
  print("x is not a number")
}
```
For the conditions themselves, we can make use of any of R's relational and logical operators. For a list of common operators, see the [appendix](#table-of-operators).
