# Conditional statements

## if statements

Conditional statements can be used when you want a piece of code to be executed only if a particular condition is met. The most basic form of these are 'if' statements. As a simple example, let's say we wanted to check if a variable `x` is less than 10. We can write something like:

```{r, purl=purl_example_code}
x <- 9

# A basic if statement
if (x < 10) {
  print("x is less than 10")
}
```

```{r, purl=purl_example_code}
x <- 11

if (x < 10) {
  print("x is less than 10")
}
```


## if...else statements

We can also specify if we want something different to happen if the condition is not met, using an 'if...else' statement:

```{r, purl=purl_example_code}
x <- 11

# A basic if...else statement
if (x < 10) {
  print("x is less than 10")
} else {
  print("x is 10 or greater")
}
```

---

Or if there are multiple conditions where we want different things to happen, we can add 'else if' commands:

```{r, purl=purl_example_code}
x <- 5

if (x < 10) {
  print("x is less than 10")
} else if (x == 10) {
  print("x is equal to 10")
} else {
  print("x is greater than 10")
}
```

---

For the conditions themselves, we can make use of R's relational and logical operators:

| Operator | Definition |
| :------: | :--------- |
|    ==    | Equal to                        |
|    !=    | Not equal to                    |
|    >     | Greater than                    |
|    <     | Less than                       |
|    >=    | Greater than or equal to        |
|    <=    | Less than or equal to           |
|  &#448;  | Or                              |
|    &     | And                             |
|    !     | Not                             |
|   %in%   | The subject appears in a vector |
| is.na()  | The subject is NA               |


## Vectorising an if...else statement

Dplyr's `if_else()` function is useful if we want to apply an 'if...else' statement to a vector, rather than a single value. When we use `if_else()` we need to provide it with three arguments, like this: `if_else(condition, true, false)`, where `condition` is the condition we want to test, `true` is the value to use if the condition evaluates to `TRUE`, and `false` is the value to use if the condition evaluates to `FALSE`.

For example, if we had a vector containing a set of numbers, and we wanted to create an equivalent vector containing a '1' if the number is greater than zero, or a '0' if the number is less than or equal to zero, then we could do:

```{r, purl=purl_example_code}
x <- c(0, 74, 0, 8, 23, 15, 3, 0, -1, 9)

# Vectorised if...else
dplyr::if_else(x > 0, 1, 0)
```

---

When we're manipulating dataframes, it can be useful to combine `if_else()` with the `mutate()` function from dplyr. Let's take a look at the `offenders` dataframe, which is also used in the [Introduction to R](https://github.com/moj-analytical-services/IntroRTraining) course:

```{r, message=F, warning=F, purl=purl_example_code}
# First read and preview the data
offenders <- Rs3tools::s3_path_to_full_df(
  "alpha-r-training/intro-r-training/Offenders_Chicago_Police_Dept_Main.csv"
)
str(offenders)
```

---

Let's say we wanted a simple way to be able to separate youths from adult offenders. We can add a column that contains 'Youth' if the offender is under the age of 18, and 'Adult' otherwise:

```{r, purl=purl_example_code}
# Now use mutate to add the new column
offenders <- offenders %>%
  dplyr::mutate(YOUTH_OR_ADULT = dplyr::if_else(AGE < 18, "Youth", "Adult"))

str(offenders)
```


## Vectorising multiple if...else statements

In the previous section we saw how we can apply a single condition to a vector, but what if we want to apply several conditions, each with a different outcome, at the same time? We can use the `case_when()` function from dplyr to do this. Let's say that we wanted to add a column to the `offenders` dataframe with an age band for each offender. We can do something like this:

```{r, purl=purl_example_code}
# Add an age band column
offenders <- offenders %>%
  dplyr::mutate(AGE_BAND = dplyr::case_when(
    AGE < 18 ~ "<18",
    AGE < 30 ~ "18-29",
    AGE < 40 ~ "30-39",
    AGE < 50 ~ "40-49",
    AGE < 60 ~ "50-59",
    AGE >= 60 ~ "60+",
    TRUE ~ "Unknown"
  ))
offenders %>% select(BIRTH_DATE, AGE, AGE_BAND) %>% str(vec.len=6)
```

---

In the `case_when()` function, each argument should be a two-sided formula. For each formula, the condition appears to the left of a '`~`' symbol, and on the right is the value to assign if the condition evaluates to `TRUE`.

Note that the order of conditional statements in the `case_when()` function is important if there are overlapping conditions. The conditions will be evaluated in the order that they appear in, so in the above example, the `case_when()` will first check if the person is under 18, then if they are under 30 (but over 18), and so on.

A default value can be assigned in the event that none of the conditions are met. This is done by putting `TRUE` in the place of a condition. In the example above, if none of the conditions are met then a value of `"Unknown"` is assigned.


---

### Exercise 1

Add a column called 'COURT_ORDER' to the `offenders` dataframe. The column should contain a '1' if the offender received a court order, or a '0' otherwise, based on the categories in the 'SENTENCE' column.

**Hint:** you'll need to apply the `if_else()` function with `mutate()`.


### Exercise 2

Add a column called 'PREV_CONVICTIONS_BAND' to the `offenders` dataframe. The column should contain the following categories: 'Low', 'Medium', 'High', based on the number of convictions given in the 'PREV_CONVICTIONS' column. For example, you can consider less than 5 PREV_CONVICTIONS to be 'Low', 5 to 10 to be 'Medium', and over 10 to be 'High'.

**Hint:** you'll need to use the `case_when()` function with `mutate()`.

---

```{r, include=show_solution, purl=purl_solutions}
# Conditional statements - solution to exercise 1
offenders <- offenders %>%
  dplyr::mutate(COURT_ORDER = dplyr::if_else(SENTENCE == "Court_order", 1, 0))

str(offenders)
```

---

```{r, include=show_solution, purl=purl_solutions}
# Conditional statements - solution to exercise 2
offenders <- offenders %>%
  dplyr::mutate(PREV_CONVICTIONS_BAND = dplyr::case_when(
    PREV_CONVICTIONS < 5 ~ "Low",
    PREV_CONVICTIONS <= 10 ~ "Medium",
    PREV_CONVICTIONS > 10 ~ "High",
    TRUE ~ "Unknown"
  ))
str(offenders)
```
