# Bonus examples

Let's take a look at a few more examples and tackle some problems that we might encounter as an analyst in DASD. 


## Example 1 - reshaping

Let's look at an example with some aggregate data based on the `offenders` dataset:

```{r, purl=purl_example_code}
offenders_summary <- offenders %>%
  group_by(REGION, SENTENCE) %>%
  summarise(offender_count = n())
offenders_summary
```

### Transforming from long to wide format

The above summary dataframe could be described as being in a 'long' format - where there are minimal columns and lots of rows. This format tends not to be used for presenting data, as it is more difficult to look at and interpret. Therefore wider formats are often used to display data, where there are more columns but fewer rows. We can use the `pivot_wider()` function from tidyr to help us transform from a long format to a wide format, like so:

```{r, purl=purl_example_code}
offenders_summary <- offenders_summary %>%
  tidyr::pivot_wider(names_from = "SENTENCE", values_from = "offender_count")

offenders_summary
```

In the `names_from` argument of `pivot_wider()`, we've specifed that we want to create new columns based on the different categories in the `SENTENCE` column - so there will be one new column for each of the three categories that appear in `SENTENCE`. Then we use the `values_from` argument to specify that we want values from the `offender_count` column to go into those new columns.


### Transforming from wide to long format

In order to reverse the reshaping that we've just done, and go back from a wide format to a long format, we can use the `pivot_longer()` function:

```{r, purl=purl_example_code}
offenders_summary <- offenders_summary %>%
  tidyr::pivot_longer(cols = -REGION, names_to = "SENTENCE", values_to = "offender_count")

offenders_summary
```

The `cols` argument of `pivot_longer()` has been set to `-REGION`, which means that all columns apart from `REGION` will be reshaped. Then the `names_to` argument is used to specify that we want the names of those columns to go into a new column called `SENTENCE`, and the `values_to` argument is used to specify that we want the values in those column to go into a new column called `offender_count`.



## Example 2

```{r, message=F, warning=F, purl=purl_example_code}
# Read data
prosecutions_and_convictions <- botor::s3_read(
  "s3://alpha-r-training/writing-functions-in-r/prosecutions-and-convictions-2018.csv", read.csv)

# Filter for Magistrates Court to extract the prosecutions
prosecutions <- prosecutions_and_convictions %>%
  filter(`Court.Type` == "Magistrates Court")
```

The following code is used to prepare a table that will form the basis of this example. This table will show the number of prosecutions over time for each offence group.

```{r, purl=purl_example_code}
# Create a time series table
time_series <- prosecutions %>%
  group_by(Year, Offence.Type, Offence.Group) %>%
  summarise(Count = sum(Count)) %>%
  # Select the past 5 years (to avoid the table being too wide)
  filter(Year > max(prosecutions$Year) - 5) %>%
  # Convert from long format to wide format
  tidyr::pivot_wider(names_from = "Year", values_from = "Count", values_fill = c("Count" = 0)) %>%
  arrange(Offence.Type, Offence.Group) %>%
  ungroup()

# This removes repeated row labels, to replicate how this data might be displayed in Excel
time_series$Offence.Type[duplicated(time_series$Offence.Type)] <- NA

time_series
```

Let's imagine we received a dataset in the above format, and we wanted to calculate the total number of prosecutions over the past 5 years for each offence type and group. In the current format, we'd need to sum the values in the columns `2013` - `2018`. We could do something like this: 

```{r, purl=purl_example_code}
total <- (time_series$`2014` + time_series$`2015` + time_series$`2016` +
          time_series$`2017` + time_series$`2018`)

time_series_with_total <- time_series
time_series_with_total$Total <- total

time_series_with_total
```

But what if we want to re-use the code in the future? We'd need to generalise it for different years or a different number of years. Fortunately we can restructure the data to help with this problem.

First let's deal with the empty row labels in the `Offence.Type` column. Although avoiding repeated row labels looks neater in an Excel table, it can be problematic for analysis. Fortunately we can easily fill the row labels in using the `fill()` function from tidyr:

```{r, purl=purl_example_code}
time_series <- time_series %>% tidyr::fill(Offence.Type)
time_series
```

Now we need to transform this dataframe into a long format, using `pivot_longer()`:

```{r, purl=purl_example_code}
time_series_long <- time_series %>%
  tidyr::pivot_longer(cols = -c("Offence.Type", "Offence.Group"), names_to = "year", values_to = "count")

time_series_long
```

Now we're ready to find the total for each offence group using `group_by()` and `summarise()` from dplyr:

```{r, purl=purl_example_code}
totals <- time_series_long %>%
  group_by(Offence.Type, Offence.Group) %>%
  summarise(Total = sum(count))

totals
```

If we wanted to add these totals to our original dataframe, we can use `left_join()` from dplyr:

```{r, purl=purl_example_code}
time_series <- dplyr::left_join(time_series, totals, by=c("Offence.Type", "Offence.Group"))

time_series
```

Now we've managed to calculate the total number of prosecutions over the past 5 years, without needing to hard-code the names of those years. This means that the code can be re-used in future years without needing to be edited.


