
# Iteration


## Introduction

'For' and 'while' loops are used to repeatedly execute a piece of code, and are are a fundamental part of most programming languages. This chapter introduces how to use them in R, as well as showing how we can iterate over the rows or columns of a dataframe without needing to write a loop.  

A general rule of thumb in programming is to avoid copying and pasting a piece of code more than once; if you find that you are repeating similar pieces of code over and over again, this is a sign that either a loop or a [function](https://github.com/moj-analytical-services/writing_functions_in_r) (or both) are required. Keeping your code concise will help make it more readable and easier to understand.


## For loop basics

Let's start with a very basic example to illustrate what a for loop does. Say we wanted to print the numbers 1 to 5; without a for loop we'd need to write something like this:

```{r, collapse = TRUE, purl=purl_example_code}
print(1)
print(2)
print(3)
print(4)
print(5)
```

Clearly there is some code repetition here, so we can achieve the same result using a for loop:

```{r, purl=purl_example_code}
for (i in 1:5) {
  print(i)
}
```

Inside the brackets of the for loop you define a variable - in this case called `i` - along with what you want to iterate over, referred to as the iterable. In this case the iterable is a sequence of the numbers 1 to 5, denoted `1:5` in R. For each iteration, the variable `i` will take on a value equal to the next element of the iterable. The loop body goes inside the curly brackets, which is where you define what you want to happen for each iteration (in this case printing the value of `i`).

In the previous example we iterated over a sequence of numbers, but in R you can iterate over anything you like. Here's a similar example, but iterating over a vector of strings instead of a sequence of numbers:

```{r, purl=purl_example_code}
strings <- c("spam", "egg", "chips", "beans")

for (string in strings) {
  print(string)
}
```

You can also use for loops to populate or modify a vector or dataframe. The following example shows how we can add the first ten numbers of the Fibonacci sequence to a vector:

```{r, purl=purl_example_code}
n <- 10 # Specify what length we want our output vector to be
fibonacci <- vector("numeric", n) # Define an empty numeric vector of length n that we'll populate using the loop

# Set up the first couple of numbers to get the sequence started
fibonacci[1] <- 0 
fibonacci[2] <- 1

# Add the rest of the sequence
for (i in 3:n) {
  fibonacci[i] <- fibonacci[i-1] + fibonacci[i-2]
}

print(fibonacci)
```

When writing a for loop you must define something to iterate over a fixed number of times in advance. It is also possible to iterate indefinitely using a different kind of loop - this is covered in the section on while loops, below.


### Exercise

Write a for loop to print "The current date is ..." for each date in the following string vector:

```{r, purl=purl_example_code}
dates <- c("2020-03-01", "2020-06-01", "2020-09-01", "2020-12-01")
```

**Hint:** You can use the `paste()` function to join strings together, and the `print()` function to print the result in the Console.

---

```{r, include=show_solution, purl=purl_solutions}
# Solution
for (date in dates) {
  print(paste("The current date is", date))
}
```

---

## More options with for loops

### Iterating over the index of a vector

If you wanted to get an index number for each element of the iterable, you can use the `seq_along()` function. For example:

```{r, purl=purl_example_code}
strings <- c("spam", "egg", "chips", "beans")

for (i in seq_along(strings)) {
  # Use paste() to combine two strings together
  print(paste(i, strings[i]))
}
```

### Handling outputs of unknown length

There are cases where the size of an output from a loop is not known beforehand. For example, this might be because different iterations result in outputs of different lengths.

Let's say we want to join together dataframes of different lengths. There is a shared folder prepared in the alpha-r-training s3 bucket, which contains some data for us to read in and combine together. First we need to get a list of file paths to read from:

```{r, message=F, warning=F, purl=purl_example_code}
# Get dataframe with all available files/folders from an s3 path
files <- botor::s3_ls("s3://alpha-r-training/intro-r-extension")
# Get a list of csv files
files <- files %>%
  filter(grepl(".csv", uri)) %>%
  pull(uri)

files
```

Now we can use a for loop to read in each file as a dataframe, and add each dataframe to a list. After the for loop, the `bind_rows()` function from dplyr can be used to combine the data into a single dataframe. This is more memory efficient than appending the dataframes in each loop iteration.

```{r, message=F, warning=F, purl=purl_example_code}
# First define an empty list
fruit_list <- list()

# Loop over each file, and add the data to a list
for (i in seq_along(files)) {
  fruit_list[[i]] <- botor::s3_read(files[i], read.csv)
}

# Combine the list of dataframes into a single dataframe
fruit <- dplyr::bind_rows(fruit_list)
fruit
```
By doing this we've combined together various segments of a dataset, without needing to know how many segments there are or how many rows are in each segment beforehand.


### Conditionally exiting a for loop

You might want to stop a for loop iterating under a certain condition. In this case you can use a `break` statement in combination with an 'if' or 'if...else' statement, like so:

```{r, purl=purl_example_code}

for (i in 1:10) {
  # Exit the for loop if i is greater than 5
  if (i > 5) {
    break
  }
  
  print(i)
}
```

### Conditionally skip to the next iteration

The `next` statement can be use to skip to the next iteration of the loop under a certain condition. For example, we can skip to the next iteration if the iterable is NA (not available):

```{r, purl=purl_example_code}

data <- c(56, 92, NA, 40, 11)

for (i in data) {
  # Skip this iteration if i is NA
  if (is.na(i)) {
    next
  }
  
  print(i)
}
```

## While loops

There may be cases where we want to keep looping over a piece of code until a certain condition is met, rather than having to specify in advance how many times a loop should run. In these cases a while loop can be used, which can be thought of as a repeating if statement.

For example, we can use a while loop to achieve a similar result to the first for loop example above:

```{r, purl=purl_example_code}
# First specify an initial value for the variable used in the while loop
i <- 1

# Now define a while loop
while (i <= 5) { # The loop will continue until the condition i<=5 is met
  print(i)
  i = i + 1 # Set the value of the variable for the next loop iteration
}
```

The syntax for a while loop is similar to that of a for loop, but in the brackets a condition is specified instead of an iterable. Prior to writing the while loop you'll also need to specify a start value for the variable used in the loop, and there should be something in the body of the loop to change the variable during each iteration - otherwise the condition can never be met!

Generally a while loop should only be used in circumstances where it isn't possible to achieve the desired result using a for loop. The reason for this is that it can be easy to accidentally set up an infinite loop, where a bug in the code means that the condition is never met for a while loop to end.


## Iterating over dataframes

Although loops are an essential programming tool, Tidyverse includes functions that allow us to apply operations across all (or a subset) of the rows and columns in a dataframe. The advantages of using these built-in functions are that they can make the code more concise and easier to read, plus they're often faster to run than the for loop equivalent.


### Iterating over rows in a dataframe

We've already met one of these functions in the [Introduction to R](https://github.com/moj-analytical-services/IntroRTraining) course: the `mutate()` function from dplyr is a convenient way to apply an operation to all rows in a single column of a dataframe. For example, going back to the `fruit` dataset that we combined together earlier, here's how we can multiply all values in the `Quantity` column by two:

```{r, purl=purl_example_code}
double_fruit <- fruit %>% dplyr::mutate(Quantity = Quantity * 2)

double_fruit
```
### Iterating over columns in a dataframe

The `across()` function from dplyr can be used with `mutate()` to allow us to apply a function to multiple columns in one go.

```{r, purl=purl_example_code}
more_fruit <- fruit %>% dplyr::mutate(fruit, across(where(is.numeric), ~ .x * 10))

more_fruit
```

