
# Iteration


## Introduction

This chapter introduces how to use loops to repeatedly execute a piece of code.

A general rule of thumb in programming is to avoid copying and pasting a piece of code more than once; if you find that you are repeating similar pieces of code over and over again, this is a sign that either a loop or a [function](https://github.com/moj-analytical-services/writing_functions_in_r) (or both) are required. Keeping your code concise will help make it more readable and easier to understand.


## For loops

Let's start with a very basic example to illustrate what a for loop does. Say we wanted to print the numbers 1 to 5; without a for loop we'd need to write something like this:

```{r, collapse = TRUE, purl=purl_example_code}
print(1)
print(2)
print(3)
print(4)
print(5)
```

Clearly there is some code repetition here, so we can achieve the same result using a for loop:

```{r, purl=purl_example_code}
for (i in 1:5) {
  print(i)
}
```

Inside the brackets of the for loop you define a variable (in this case called `i`), along with what you want to iterate over (referred to as the iterable - in this case the numbers 1 to 5, denoted `1:5` in R). For each iteration, the variable `i` will take on a value equal to the next element of the iterable. The loop body goes inside the curly brackets, which is where you define what you want to happen for each iteration (in this case printing the value of `i`).

In the previous example we iterated over a sequence of numbers, but in R you can iterate over anything you like. Here's a similar example, but iterating over a vector of strings instead of a sequence of numbers:

```{r, purl=purl_example_code}
strings <- c("spam", "egg", "chips", "beans")

for (string in strings) {
  print(string)
}
```

If you wanted to get an index number for each element of the iterable, you can use the `seq_along()` function. For example:

```{r, purl=purl_example_code}
strings <- c("spam", "egg", "chips", "beans")

for (i in seq_along(strings)) {
  # Use paste() to combine two strings together
  print(paste(i, strings[i]))
}
```

You can also use for loops to populate or modify a vector or dataframe. The following example shows how we can add the first ten numbers of the Fibonacci sequence to a vector:

```{r, purl=purl_example_code}
n <- 10 # Specify what length we want our output vector to be
fibo <- vector("numeric", n) # Define an empty numeric vector of length n that we'll populate using the loop

# Set up the first couple of numbers to get the sequence started
fibo[1] <- 0 
fibo[2] <- 1

# Add the rest of the sequence
for (i in 3:n) {
  fibo[i] <- fibo[i-1] + fibo[i-2]
}

print(fibo)
```

When writing a for loop you must define something to iterate over a fixed number of times in advance. It is also possible to iterate indefinitely using a different kind of loop - this is covered in the section on while loops, below.


### Exercise

Write a for loop to print "The current date is ..." for each date in the following string vector:

```{r, purl=purl_example_code}
dates <- c("2020-03-01", "2020-06-01", "2020-09-01", "2020-12-01")
```

**Hint:** You can use the `paste()` function to join strings together.

---
```{r, include=show_solution, purl=purl_solutions}
# Solution
for (date in dates) {
  print(paste("The current date is", date))
}
```
---


### Handling outputs of unknown length

There are cases where the size of an output from a loop is not known beforehand. For example, this might be because different iterations result in outputs of different lengths.

Let's say we want to join together dataframes of different lengths. There is a shared folder prepared in the alpha-r-training s3 bucket, which contains some data for us to read in and combine together. First we need to get a list of file paths to read from:

```{r, message=F, warning=F, purl=purl_example_code}
# Get a vector of file paths to iterate over
files <- s3tools::list_files_in_buckets("alpha-r-training", prefix="intro-r-extension", path_only=T)
files <- files[grepl(".csv", files)] # This makes sure that only csv files are included

files
```

Now we can use a for loop to read in each file as a dataframe, and add each dataframe to a list. After the for loop, the `bind_rows()` function from dplyr can be used to combine the data into a single dataframe. This is more memory efficient than appending the dataframes in each loop iteration.

```{r, message=F, warning=F, purl=purl_example_code}
# First define an empty list with the same length as the number of datasets to be read in
fruit_list <- vector("list", length(files))

# Loop over each file, and add the data to a list
for (i in seq_along(files)) {
  fruit_list[[i]] <- s3tools::s3_path_to_full_df(files[i])
}

# Combine the list of dataframes into a single dataframe
fruit <- dplyr::bind_rows(fruit_list)
fruit
```

### Conditionally exiting a for loop

You might want to stop a for loop iterating under a certain condition. In this case you can use a `break` statement in combination with an 'if' or 'if...else' statement, like so:

```{r, purl=purl_example_code}

for (i in 1:10) {
  # Exit the for loop if i is greater than 5
  if (i > 5){
    break
  }
  
  print(i)
}
```

### Conditionally skip to the next iteration

The `next` statement can be use to skip to the next iteration of the loop under a certain condition. For example, we can skip to the next iteration if the iterable is equal to 3:

```{r, purl=purl_example_code}

for (i in 1:5) {
  # Skip this iteration if i equals 3
  if (i == 3){
    next
  }
  
  print(i)
}
```


## While loops

There may be cases where we want to keep looping over a piece of code until a certain condition is met, rather than having to specify in advance how many times a loop should run. In these cases a while loop can be used.

For example, we can use a while loop to achieve a similar result to the first for loop example above:

```{r, purl=purl_example_code}
# First specify an initial value for the variable used in the while loop
i <- 1

# Now define a while loop
while (i <= 5) { # The loop will continue until the condition i<=5 is met
  print(i)
  i = i + 1 # Set the value of the variable for the next loop iteration
}
```

The syntax for a while loop is similar to that of a for loop, but in the brackets a condition is specified instead of an iterable. Prior to writing the while loop you'll also need to specify a start value for the variable used in the loop, and there should be something in the body of the loop to change the variable during each iteration - otherwise the condition can never be met!

Generally a while loop should only be used in circumstances where it isn't possible to achieve the desired result using a for loop. The reason for this is that it's quite easy to accidentally set up an infinite loop, where a bug in the code means that the condition is never met for a while loop to end.

