# Handling missing data

It's often the case that datasets will contain missing values, which are usually denoted by `NA` in R. 'NA' stands for 'not available', while other programming languages might use 'NaN' (not a number) or 'null' instead. Care needs to be taken to make sure these missing values are handled in the most appropriate way for a particular situation. This section introduces a few methods for handling missing values in different situations.


## Identifying missing values

The function `is.na()` can be used to identify missing values, and it can be applied to a single value or a vector. It returns `TRUE` if a value is `NA`, and `FALSE` otherwise:

```{r, purl=purl_example_code}
x <- 7
is.na(x)

x <- c(7, 23, 5, 14, NA, 1, 11, NA)
is.na(x)
```

If instead you wanted to identify values that are not missing, then you can combine `is.na()` with the 'not' operator, `!`, like so:
```{r, purl=purl_example_code}
x <- c(7, 23, 5, 14, NA, 1, 11, NA)
!is.na(x)
```

When working with dataframes, the `complete.cases()` function is useful to check which rows are complete (i.e. the row doesn't contain any missing values):
```{r, purl=purl_example_code}
df <- tibble::tibble(
  "x" = c(0, 1, 2, NA, 4),
  "y" = c(18, NA, 45, 15, 2),
)

complete.cases(df)
```




## Handling missing values using a function argument

Some functions have built-in arguments where you can specify what you want to happen to missing values. For example, the `sum()` function has an argument called `na.rm` that you can use to specify if you want the `NA` values to be removed before the sum is calculated. By default `na.rm` is set to `FALSE`:

```{r, purl=purl_example_code}
x <- c(7, 23, 5, 14, NA, 1, 11, NA)
sum(x)
```

So if we try to sum over numeric vector that contains `NA` values, then the result is `NA`. By setting `na.rm` to `TRUE`, however, we can remove the `NA` values before continuing with the sum:

```{r, purl=purl_example_code}
x <- c(7, 23, 5, 14, NA, 1, 11, NA)
sum(x, na.rm=TRUE)
```


## Converting values to NA

There might be occasions where we want to set some values to `NA`, for example if they are invalid. The `replace()` function can be used to replace values based on a particular condition. This example shows how negative values in a vector can be replaced with `NA`:

```{r, purl=purl_example_code}
x <- c(7, 23, 5, -14, 0, -1, 11, 0)
replace(x, x < 0, NA)
```


## Replacing missing values

Sometimes it's necessary to replace missing values; for example, when displaying data in a table or chart, or when the missing values would cause problems for a particular calculation. There are a few different options that we'll visit in this section.


### Replacing with a specific value

We can also use the `replace()` function to replace missing values with a specific value. In this example we're replacing missing values with zero:

```{r, purl=purl_example_code}
x <- c(7, 23, 5, -14, NA, -1, 11,NA)
replace(x, is.na(x), 0)
```

The `replace()` function can also be applied to a whole dataframe, like so:

```{r, purl=purl_example_code}
df <- tibble::data_frame(
  "x" = c(0, 1, 2, NA, 4),
  "y" = c(18, NA, 45, 15, 2),
)

df %>% replace(is.na(.), 0)
```

The `replace_na()` function from tidyr also provides a convenient way to replace missing values in a dataframe, and is especially useful if you want to use different replacement values for different columns.

Here's an example of how to use `replace_na()` with the `offenders` dataframe, where we're replacing missing values in the `HEIGHT` column with 'Unknown':
```{r, purl=purl_example_code}
offenders_replacena <- offenders %>%
  dplyr::mutate(HEIGHT = as.character(HEIGHT)) %>%
  tidyr::replace_na(list(HEIGHT = "Unknown"))

# Display the dataframe in descending height order, so we can see the 'Unknown' values
offenders_replacena %>% dplyr::arrange(desc(HEIGHT)) %>% str()
```



---

### Replacing with values from another column

The `coalesce()` function from dplyr can be used to fill in missing values with values from another column. Before we jump into an example, let's first prepare a dataframe:

```{r, purl=purl_example_code, message=F, warning=F}
event_dates <- tibble::tibble(
  "event_id" = c(0, 1, 2, 3, 4, 5),
  "date" = c("2016-04-13", "2015-12-29", "2016-06-02", "2017-01-27", "2015-10-21", "2018-03-15"),
  "new_date" = c("2016-08-16", NA, NA, "2017-03-02", NA, "2018-11-20")
)

event_dates
```

We can use `coalesce()` to fill in the missing dates in the `new_date` column with the equivalent date in the `date` column:

```{r, purl=purl_example_code}
event_dates %>%
  dplyr::mutate(new_date = dplyr::coalesce(new_date, date))
```

### Replacing with the previous value in a column

If we were to encounter a dataframe like the following, where each group name in the `year` column appears only once, then we might want to fill in the missing labels before beginning any analysis.

```{r, purl=purl_example_code}
# Construct the example dataframe
df <- tidyr::crossing(year = c("2017", "2018", "2019"),
                      quarter = c("Q1", "Q2", "Q3", "Q4")) %>%
      dplyr::mutate(count = sample(length(year)))

# This removes repeated row labels
df$year[duplicated(df$year)] <- NA

df
```

The `fill()` function from tidyr is a convenient way to do this, and can be used like this:

```{r, purl=purl_example_code}
df %>% tidyr::fill(year)
```

## Removing rows with missing values from a dataframe

The `drop_na()` function from tidyr allows us to easily remove rows containing `NA` values from a dataframe. Let's say we wanted to remove all incomplete rows from the `offenders` dataset. We can either do this:
```{r, purl=purl_example_code}
offenders_nona <- offenders %>% tidyr::drop_na()

str(offenders_nona)
```
Or alternatively you can remove rows that contain `NA` values in specific columns:
```{r, purl=purl_example_code}
offenders_nona <- offenders %>% tidyr::drop_na(HEIGHT, WEIGHT)

str(offenders_nona)
```


---

### Exercise

For the following dataframe, use the `filter()` function from dplyr with `complete.cases()` to extract the rows **with** missing values:
```{r, purl=purl_example_code}
fruit <- tibble::tibble(
  "Item" = c("Orange", "Apple", "Banana", "Lemon", "Pear"),
  "Cost" = c(0.5, 0.4, 0.1, 0.3, NA),
  "Quantity" = c(23, NA, 15, 9, 11)
)
```
**Hint:** you can use a `.` inside the `complete.cases()` function to apply it to all columns of the dataframe.

---

```{r, include=show_solution, purl=purl_solutions}
# Solution
fruit %>% dplyr::filter(!complete.cases(.))
```

---

### Exercise

For the following dataframe, use the `replace_na()` function from tidyr to replace missing values in the `Cost` column with "Unknown" and the `Quantity` column with 0.
```{r, purl=purl_example_code}
fruit <- tibble::tibble(
  "Item" = c("Orange", "Apple", "Banana", "Lemon", "Pear"),
  "Cost" = c("£0.50", "£0.40", "£0.10", "£0.30", NA),
  "Quantity" = c(23, NA, 15, 9, 11)
)
```
**Hint:** you can add multiple arguments to `replace_na(list(...))`, with one argument for each column where NA values need replacing.

---

```{r, include=show_solution, purl=purl_solutions}
# Solution
fruit %>% tidyr::replace_na(list(Cost = "Unknown",
                                 Quantity = 0))
```



