---
title: Introduction to R extension
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- FALSE # This determines if the solutions are displayed in the readme
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions

# Change botor settings to avoid printing debugger messages to the console
logger::log_threshold('WARN', namespace = 'botor')
```

This repository is for the Introduction to R+ course offered by the Data & Analysis R Training Group. 

The session is intended to be accessible to anyone who is familiar with the content of the [Introduction to R](https://github.com/moj-analytical-services/IntroRTraining) training session.

If you have any feedback on the content, please get in touch!

## Contents

* [Pre-material](#pre-material)
* [Learning outcomes](#learning-outcomes)
* [Conditional statements](#conditional-statements)
* [Handling missing data](#handling-missing-data)
* [Iteration](#iteration)
* [Reshaping data](#reshaping-data)
* [String manipulation](#string-manipulation)
* [Further reading](#further-reading)
* [Bonus examples](#bonus-examples)
* [Appendix](#appendix)


## Pre-material

Before the session, please make sure that -

1. You have access to RStudio on the Analytical Platform
2. You have access to the [alpha-r-training s3 bucket](https://controlpanel.services.analytical-platform.service.justice.gov.uk/datasources/607/)
3. You have followed the steps in the [Configure Git and Github section of the Platform User Guidance](https://user-guidance.services.alpha.mojanalytics.xyz/github.html#setup-github-keys-to-access-it-from-r-studio-and-jupyter) to configure Git and GitHub (this only needs doing once)
4. You have cloned this repository (instructions are in the Analytical Platform User Guidance if you follow step 1 [here](https://user-guidance.services.alpha.mojanalytics.xyz/github.html#r-studio))
5. You have installed the required packages by entering the following commands in the Console window in RStudio (after following step 4, above): `install.packages("renv")` followed by `renv::restore()`

If you have any problems with the above please get in touch with the course organisers or ask for help on either the #analytical-platform-support or #intro_r channel on [ASD slack](https://asdslack.slack.com).

All the examples in the presentation and README are available in the R script example_code.R.


```{r source_content, child = 'introduction.Rmd'}
```

```{r source_content, child = 'conditional_statements.Rmd'}
```

```{r source_content, child = 'iteration.Rmd'}
```

```{r source_content, child = 'missing_data.Rmd'}
```

```{r source_content, child = 'reshaping.Rmd'}
```

```{r source_content, child = 'strings.Rmd'}
```

```{r source_content, child = 'further_reading.Rmd'}
```

```{r source_content, child = 'examples.Rmd'}
```

```{r source_content, child = 'appendix.Rmd'}
```
