---
title: Introduction to R extension
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- TRUE # This determines if the solutions are displayed in the slides
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions

# Change botor settings to avoid printing debugger messages to the console
logger::log_threshold('WARN', namespace = 'botor')
```

## Setup

For those of you who'd like to 'code along' and attempt the exercises, please make sure you've:

1. Got access to the `alpha-r-training` s3 bucket
2. Cloned the [intro_r_training_extension repo](https://github.com/moj-analytical-services/intro_r_training_extension) (or copied example_code.R into your RStudio) - instructions on cloning from GitHub can be found [here](https://user-guidance.services.alpha.mojanalytics.xyz/github.html#r-studio)
3. Installed the required packages by entering the following commands in the Console window in RStudio: `install.packages("renv")` followed by `renv::restore()`
4. Opened `example_code.R` in RStudio
5. Opened the [intro_r_training_extension repo](https://github.com/moj-analytical-services/intro_r_training_extension) on GitHub, so you can refer to the session material in the README

If you need any help, just ask! We will be monitoring Teams throughout the session.


```{r source_content, child = 'introduction.Rmd'}
```

```{r source_content, child = 'conditional_statements.Rmd'}
```

```{r source_content, child = 'iteration.Rmd'}
```

```{r source_content, child = 'missing_data.Rmd'}
```

```{r source_content, child = 'reshaping.Rmd'}
```

```{r source_content, child = 'strings.Rmd'}
```

```{r source_content, child = 'further_reading.Rmd'}
```


# Any questions?

