---
title: Introduction to R extension
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- TRUE # This determines if the solutions are displayed in the slides
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
```

## Introduction to R: setup

Before we start, make sure you've:

1. Got access to the `alpha-r-training` bucket
2. Cloned the `intro_r_training_extension` repo
3. Pulled any changes (the blue arrow on the Git tab in the top right panel)
4. Got `example_code.R` open
5. Opened the `intro_r_training_extension` repo on GitHub, so you can refer to the read me

If you need any help, just ask! For anyone joining remotely, we will be monitoring Teams throughout the session.


```{r source_content, child = 'introduction.Rmd'}
```

```{r source_content, child = 'conditional_statements.Rmd'}
```

```{r source_content, child = 'iteration.Rmd'}
```

```{r source_content, child = 'reshaping.Rmd'}
```

```{r source_content, child = 'strings.Rmd'}
```

# Further material

Go to https://github.com/moj-analytical-services/intro_r_training_extension and take a look at the README to find more advanced examples and further reading.


# Any questions?
