# String manipulation


```{r, echo=F, message=F, results=F}

# loading packages
# 
library(tidyverse)
library(stringr)

```

## Introduction

In this part we will introduce strings and how to manipulate them to suit the needs of a particular project. The packages needed (listed above) are mostly the `stringr` and  `tidyverse` package where the former contains a great number of functions to manipulate and work with strings and the latter is the base for most of the things we do in R. 

It should also be mentioned that the majority of the content that follows is containe din the book `R for Data Science` by `Hadley Wickam` and `Garrett Grolemund`.

## String Basics

There are two ways with which you can create a string in R, at the very basic level, by either using single of double quotes. There is no difference in the behaviour but it is recomended to stick with `"` as it is more widely understood that this is a string and others will be able to follow your code esier. 

To note here that one of the advantages of having two ways to define a string is that you can combine them when quotes within a string if needed as showcased in the example below. 


```{r}
string1 = "a string using double quotes"

is.character(string1)

string2 = 'another string using single quotes'

is.character(string2)

# a string containing quotes
string3 = "this is a 'string' within a string"

# notice how the output changes when implementing the following code
string4 = 'this is a "string" within a string'


```


Notice the difference in `string4`. This happens becaus the compiler cannot place two double quotes within the same string and so we need to `escape` it. To achieve that a `\` is used before the corresponding character. This can be used not only for `"` but for any other reserved character as well. 

If you need to `escape` a backslash then use a double backslash as in `\\` and to see what the output would look like when used in text you can use the `writeLines()` function to simulate that effect. 


```{r}
string5 = "escaping a reserved character like \" quotes "

# to see how the result would apear in text
writeLines(string5)

string6 = "escaping a backslash \\ "

# to see how the result would apear in text
writeLines(string6)

```



Fuirthermore, additional control in the form of newlines or tabs is provided by using `\n` and `\t` for example along with a plethora of available special characters that can be found by typing `?"'"` or `?'"'` in you command line. 

Introducing non-English characters in your text strings is also possible using special cahracters of the form `\u00b5` for example for the greek letter mu. 

```{r}

# outputting non-English characters
string7 = "\u00b5" 


```


Adding strings into a character vector is very common and is usualy done with the use of the `c()` function. 

```{r}

s8 = c("a", "vector", "of", "strings")


```



## String Length



## Combining Strings

## Subseting strings

## Customising text for different language locales

## Exercises


## Regular Expressions
















