# String manipulation


### Introduction

In this chapter we look at strings and how to work with them. The `stringr` package from `tidyverse` contains a great number of functions to help manipulate strings. 

There are two ways to create a string in R, by using either single or double quotes. There is no practical difference in behaviour for the two options, but the convention is to use double quotes (`"`). 

There is an advantage to having two ways to define a string, which is that the two quotation marks can be combined for cases when the string itself needs to contain a quotation mark. Here are some examples of how to define a string in R: 

```{r, purl=purl_example_code}
string1 <- "a string using double quotes"
string2 <- 'another string using single quotes'

string1
string2

# Some strings containing quotation marks
string3 <- "here is a 'quote' within a string"
string4 <- 'here is a "quote" within a string'

string3
string4
```

Notice the difference in `string4` - R has added escape characters (`\`) before the double quote marks. These escape characters change the behaviour of the following character. In this case it stops the following double quote mark from defining the end of the string, and instead allows it to be a part of the string.

Often it is necessary to work with a set of strings in a character vector, and in the following sections we'll look at how various `stringr` functions can help us work with character vectors. A new character vector can be constructed using the `c()` function:

```{r, purl=purl_example_code}
string_vector <- c("a", "vector", "of", "strings")
```

### String length

The first `stringr` function we'll look at is `str_length()`, which returns the length of each string in a character vector:

```{r, purl=purl_example_code}
# Find out how many characters are in each string
stringr::str_length(string_vector)
```

### Combining strings

The `str_c()` function is used to combine multiple strings together, like so:

```{r, purl=purl_example_code}
stringr::str_c("some", "strings", "to", "combine")
```

There are two optional arguments, `sep` and `collapse` that can be used to modify the behaviour of `str_c()`. The `sep` argument allows us to define a separator to put between the strings when they're combined:

```{r, purl=purl_example_code}
# Using custom separator
stringr::str_c("some", "space", "separated", "strings", sep = " ")
```

The `str_c()` is especially useful because it is vectorised, and when applying it to character vectors the `collapse` argument can be used to combine a vector of strings into a single string:

```{r, purl=purl_example_code}
# Collasping a character vector
vector_to_collapse <- c("some", "strings", "to", "combine")
stringr::str_c(vector_to_collapse, collapse="")
```
The value of `collapse` will determine how the collapsed strings are separated.

You can input multiple string vectors to `str_c()`, and it will combine them together:

```{r, purl=purl_example_code}
string_vector1 <- c("A", "B", "C", "D")
string_vector2 <- c("1", "2", "3", "4")
stringr::str_c(string_vector1, string_vector2, sep=" - ")
```

```{r, purl=purl_example_code}
string_vector1 <- c("A", "B", "C", "D")
string_vector2 <- c("1", "2", "3", "4")
stringr::str_c(string_vector1, string_vector2, sep=" - ", collapse=" ")
```

It can also combine a single string with a vector of strings, like so: 

```{r, purl=purl_example_code}
# The single string will be 'recycled' to match the length of the vector
stringr::str_c("a", c("b", "c", "d"), sep = " ")

# Compare with what happens when we combine these strings with c() 
c("a", c("b", "c", "d"))
```

### Subsetting strings

Selecting part of a string can be done using the `str_sub()` function. The `start` and `end` arguments are used to define the part of the string you want to extract.

```{r, purl=purl_example_code}
x <- c("First value", "Second value", "Third value")
stringr::str_sub(x, start=1, end=3)

# Negative values for the start and end count backwards from the end of the string
stringr::str_sub(x, start=-5, end=-1)
```
Note that the `str_sub()` will still work if the end point is beyond the length of the string.


### Detecting a matched pattern

The `str_detect()` function can be used to check if part of a string matches a particular pattern. For example, let's say we wanted to check if any strings in a string vector contain "blue":   

```{r, purl=purl_example_code}
colours <- c("scarlet red", "ultramarine blue", "cadmium red", "cobalt blue", "cerulean blue")
stringr::str_detect(colours, "blue")
```

Because booleans (`TRUE` and `FALSE`) can be represented as numbers (1 and 0), you can apply numeric functions to boolean vectors:

```{r, purl=purl_example_code}
# Count how many strings contain 'blue'
sum(stringr::str_detect(colours, "blue"))
```

#### Regular expressions

Regular expressions (regex) are extremely helpful for pattern matching. Since regex could be an entire course by itself, here we only introduce a few basics to get started. See the [further reading](#further-reading) section if you're interested in learning more.

* `[:alpha:]` or `[A-Za-z]`: All uppercase and lowercase letters
* `[:digit:]` or `[0-9]`: All numbers
* `[:space:]` or `\\s`: A single space
* `^a`: Begins with 'a'
* `a$`: Ends with 'a'
* `[^a]`: Anything other than 'a'

We can use regex to help extract a more general pattern, such as only strings that contain letters:
```{r, purl=purl_example_code}
colours <- c("1.", "ultramarine blue", "2. cadmium red", "cobalt blue", "-")
stringr::str_detect(colours, "[:alpha:]")
```
Or only strings that contain letters and numbers:
```{r, purl=purl_example_code}
stringr::str_detect(colours, "[:alnum:]")
```
Or only strings that contain something other than letters, numbers, and spaces:
```{r, purl=purl_example_code}
stringr::str_detect(colours, "[^[:alnum:][:space:]]")
```


### Extracting a matched pattern

We can use the `str_extract()` function to extract strings that match a particular pattern:

```{r, purl=purl_example_code}
colours <- c("scarlet red", "ultramarine blue", "cadmium red", "cobalt blue", "cerulean blue")
stringr::str_extract(colours, "blue")
```


### Replacing a matched pattern

You can use the `str_replace()` and `str_replace_all()` functions to find and replace parts of a string. `str_replace()` replaces the first instance of the pattern, whereas `str_replace_all()` replaces all instances of the pattern:

```{r, purl=purl_example_code}
colours <- c("scarlet...red", "ultramarine.blue", "cadmium_red", "cobalt blue", "cerulean-blue")

# Replace the first character that isn't a letter or number with an underscore
stringr::str_replace(colours, "[^[:alnum:]]", "_")
# Replace all characters that aren't a letter or number with an underscore
stringr::str_replace_all(colours, "[^[:alnum:]]", "_")
```


### Exercise

Remove all spaces from the following string:

```{r, purl=purl_example_code}
string <- "The quick brown fox jumps over the lazy dog."
```

**Hint:** You can remove a matched pattern by replacing it with an empty string (`""`).

---

```{r, include=show_solution, purl=purl_solutions}
# Solution
string <- stringr::str_replace_all(string, " ", "")
string
```

---
