# String manipulation


## Introduction

In this chapter we'll look at strings and some techniques to help work with them, mainly making use of the `stringr` package from Tidyverse.

There are two ways to create a string in R, by using either single or double quotes. There is no practical difference in behaviour for the two options, but the convention is to use double quotes (`"`). 

```{r, purl=purl_example_code}
# Two options to define a string
string1 <- "a string using double quotes"
string2 <- 'another string using single quotes'

string1
string2
```

---

There is an advantage to having two ways to define a string, which is that the two types of quotation marks can be combined for cases when the string itself needs to contain a quotation mark. Here are some examples of how to define a string in R: 

```{r, purl=purl_example_code}
# Some strings containing quotation marks
string3 <- "here is a 'quote' within a string"
string4 <- 'here is a "quote" within a string'

string3
string4
```

Notice the difference in how `string4` is displayed - R has added escape characters (`\`) before the double quote marks. These escape characters change the behaviour of the following character. In this case it stops the following double quote mark from defining the end of the string, and instead allows it to be a part of the string.

---

Often it is necessary to work with a set of strings in a character vector, and in the following sections we'll look at how various `stringr` functions can help us work with character vectors. A new character vector can be constructed using the `c()` function that we've met before:

```{r, purl=purl_example_code}
string_vector <- c("a", "vector", "of", "strings")
string_vector
```


## String Length

The first `stringr` function we'll look at is `str_length()`, which simply returns the length of each string in a character vector:

```{r, purl=purl_example_code}
# Find out how many characters are in each string
stringr::str_length(string_vector)
```


## Combining Strings

The `str_c()` function is used to combine multiple strings together, where each string is included as a separate argument, like so:

```{r, purl=purl_example_code}
stringr::str_c("some", "strings", "to", "combine")
```

There are two optional arguments, `sep` and `collapse` that can be used to modify the behaviour of `str_c()`. The `sep` argument allows us to define a separator to put between the strings when they're combined:

```{r, purl=purl_example_code}
# Using custom separator
stringr::str_c("some", "space", "separated", "strings", sep = " ")
```

The `str_c()` is especially useful because it is vectorised, and when applying it to character vectors the `collapse` argument can be used to combine a vector of strings into a single string:

```{r, purl=purl_example_code}
# Collasping a character vector
vector_to_collapse <- c("some", "strings", "to", "combine")
stringr::str_c(vector_to_collapse, collapse="")
```

The value of `collapse` will determine how the collapsed strings are separated.

---

You can input multiple character vectors to `str_c()` and it will combine them together. You can either output another character vector:

```{r, purl=purl_example_code}
string_vector1 <- c("A", "B", "C", "D")
string_vector2 <- c("1", "2", "3", "4")
stringr::str_c(string_vector1, string_vector2, sep=" - ")
```

Or collapse the vectors into a single string:

```{r, purl=purl_example_code}
string_vector1 <- c("A", "B", "C", "D")
string_vector2 <- c("1", "2", "3", "4")
stringr::str_c(string_vector1, string_vector2, sep=" - ", collapse=" ")
```

It can also combine a single string with a vector of strings, like so: 

```{r, purl=purl_example_code}
# The single string will be 'recycled' to match the length of the vector
stringr::str_c("a", c("b", "c", "d"), sep = " ")

# Compare with what happens when we combine these strings with c() 
c("a", c("b", "c", "d"))
```

It's worth noting what happens if you pass vectors of different lengths to `str_c()`:

```{r, purl=purl_example_code}
string_vector1 <- c("A", "B", "C")
string_vector2 <- c("1", "2", "3", "4", "5")
stringr::str_c(string_vector1, string_vector2, sep=" - ")
```

The code produces a warning but otherwise runs. In the output, you can see that the elements of the shorter vector have been repeated when combined with the additional elements of the longer vector.


## Extracing and replacing substrings

Selecting part of a string can be done using the `str_sub()` function. The `start` and `end` arguments are used to define the part of the string you want to extract.

```{r, purl=purl_example_code}
x <- c("First value", "Second value", "Third value")
stringr::str_sub(x, start=1, end=3)

# Negative values for the start and end count backwards from the end of the string
stringr::str_sub(x, start=-5, end=-1)
```

You can also use `str_sub()` to help replace substrings:

```{r, purl=purl_example_code}
stringr::str_sub(x, start=-5, end=-1) <- "item"
x
```

## Detecting a matched pattern

The `str_detect()` function can be used to check if part of a string matches a particular pattern. For example, let's say we wanted to check if any strings in a characters vector contain "blue":   

```{r, purl=purl_example_code}
colours <- c("scarlet red", "ultramarine blue", "cadmium red", "cobalt blue", "cerulean blue")
stringr::str_detect(colours, "blue")
```

Because booleans (`TRUE` or `FALSE`) can be represented as numbers (1 or 0), you can apply some functions typically used for numbers to boolean vectors:

```{r, purl=purl_example_code}
# Count how many strings contain 'blue'
sum(stringr::str_detect(colours, "blue"))
```

There can be unintended consequences for pattern matching, let's say we wanted to find strings containing the colour "red" in another character vector:

```{r, purl=purl_example_code}
colours <- c("scarlet red", "ultramarine blue", "cadmium red", "cobalt blue", "weathered")
stringr::str_detect(colours, "red")
```

There are options to help deal with cases like this that we'll visit later on.

---

### Regular expressions

Regular expressions (regex) are extremely helpful for pattern matching. Since regex could be an entire course by itself, here we only introduce a few basics to get started. See the [further reading](#further-reading) section if you're interested in learning more about regex.

There's a common syntax for defining the patterns to match that can be used across multiple programming languages. Here are a few patterns to get started with:

* `[A-Za-z]`: All uppercase and lowercase letters
* `[0-9]`: All numbers
* `[A-Za-z0-9]`: All letters and all numbers
* `\\s`: A single space
* `^a`: Begins with 'a'
* `a$`: Ends with 'a'
* `[^a]`: Anything other than 'a'
* `\\b`: A word boundary (e.g. a space, punctuation mark or the start/end of a string)

R also contains some pre-built regex classes that you might also encounter, for example `[:alpha:]` to match all letters and `[:digit:]` to match all numbers.

---

We can use regex to help extract a more general pattern, such as only strings that contain letters:

```{r, purl=purl_example_code}
colours <- c("1.", "ultramarine blue", "2. cadmium red", "cobalt blue", "-")
stringr::str_detect(colours, "[A-Za-z]")
```

Or only strings that contain letters and numbers:

```{r, purl=purl_example_code}
stringr::str_detect(colours, "[A-Za-z0-9]")
```

Or only strings that contain something other than letters, numbers, and spaces:

```{r, purl=purl_example_code}
stringr::str_detect(colours, "[^[A-Za-z0-9\\s]]")
```

---

We can revisit the example from earlier, where we wanted to identify strings containing the colour 'red':

```{r, purl=purl_example_code}
colours <- c("scarlet red", "ultramarine blue", "cadmium red", "cobalt blue", "weathered")
stringr::str_detect(colours, "\\bred\\b")
```

The word boundary regex allows us to exclude words like "weathered" when looking for the word "red".

---

## Extracting a matched pattern

We can use the `str_extract()` function to extract strings that match a particular pattern:

```{r, purl=purl_example_code}
colours <- c("scarlet red", "ultramarine blue", "cadmium red", "cobalt blue", "cerulean blue")
stringr::str_extract(colours, "blue")
```


## Replacing a matched pattern

You can use the `str_replace()` and `str_replace_all()` functions to find and replace parts of a string. `str_replace()` replaces the first instance of the pattern, whereas `str_replace_all()` replaces all instances of the pattern. Here's an alternative version of an example we saw earlier, using a different approach to replace "value" with "item":

```{r, purl=purl_example_code}
x <- c("First value", "Second value", "Third value")
# Replace 'value' with 'item'
stringr::str_replace(x, "value", "item")
```

---

Regular expressions are also useful for string replacement. Here's a example that replaces characters that aren't letters or numbers with an underscore:

```{r, purl=purl_example_code}
colours <- c("scarlet...red", "ultramarine.blue", "cadmium_red", "cobalt blue", "cerulean-blue")

# Replace the first character that isn't a letter or number with an underscore
stringr::str_replace(colours, "[^[A-Za-z0-9]]", "_")
# Replace all characters that aren't a letter or number with an underscore
stringr::str_replace_all(colours, "[^[A-Za-z0-9]]", "_")
```


### Exercise 1

Remove all spaces from the following string:

```{r, purl=purl_example_code}
string <- "The quick brown fox jumps over the lazy dog."
```

**Hint:** You can remove a matched pattern by replacing it with an empty string (`""`).

---

```{r, include=show_solution, purl=purl_solutions}
# Solution
string <- stringr::str_replace_all(string, " ", "")
string
```

---

