---
title: Introduction to conditional statements and loops in R
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE, purl=FALSE}
# These variables determine whether or not exercise solutions are included
# (if running from render.R then these will already be defined)
if (!(exists("show_solution") & exists("purl_solutions") & exists("purl_example_code"))) {
  show_solution <- TRUE # This determines if the solutions are displayed in the slides
  purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
  purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
}


# Change botor settings to avoid printing debugger messages to the console
logger::log_threshold('WARN', namespace = 'botor')
```

## Setup

For those of you who'd like to 'code along' and attempt the exercises, please make sure you've:

1. Got access to the `alpha-r-training` bucket
2. Cloned the [C&C repo](https://github.com/moj-analytical-services/Coffee-and-Coding) (or copied example_code.R into your RStudio) - instructions on cloning from GitHub can be found [here](https://user-guidance.services.alpha.mojanalytics.xyz/github.html#r-studio)
3. Installed the packages `tidyverse` and `botor` - see guidance on how to do this [here](https://user-guidance.services.alpha.mojanalytics.xyz/appendix/botor.html#installation)
4. Opened `example_code.R` in RStudio
5. Opened the [C&C repo](https://github.com/moj-analytical-services/Coffee-and-Coding/tree/master/2022-03-09 Introduction to conditional statements and loops in R) on GitHub, so you can refer to the session material in the README



## Introduction

This Coffee and Coding session is trialling some new material that will build on the original [Introduction to R training course](https://github.com/moj-analytical-services/IntroRTraining), therefore we appreciate any feedback. It provides examples that demonstrate how the tidyverse packages can assist with tasks typically encountered in MoJ Data & Analysis.


### By the end of this session you should know how to:

 - Change what the code does based on a condition
 - Classify values in a dataframe, based on a set of conditions
 - Read and combine data from multiple csv files
 - Easily apply a function to multiple columns in a dataframe


## Before we start

To follow along with the code and participate in the exercises, open the script "example_code.R" in RStudio. All the code that we'll show in this session is stored in "example_code.R", and you can edit this script to write solutions to the exercises. You may also want to have the [session material](https://github.com/moj-analytical-services/Coffee-and-Coding/tree/master/2022-03-09 Introduction to conditional statements and loops in R) open as a reference.

First, we need to load a few packages:

```{r message=F, warning=F, purl=purl_example_code}
# Load packages
library(botor) # Used to help R interact with s3 cloud storage
library(dplyr) # Used for data manipulation
library(readr) # Used to help read in data
```


```{r source_content, child = '../rmd_files/conditional_statements.Rmd'}
```

```{r source_content, child = '../rmd_files/iteration.Rmd'}
```


# Further Reading

## Further Reading

### General

 * [Tidyverse website](https://www.tidyverse.org)
 * [MoJ Analytical Platform Guidance](https://user-guidance.services.alpha.mojanalytics.xyz)
 * [MoJ coding standards](https://moj-analytical-services.github.io/our-coding-standards/)
 
### Conditional statements

 * [Advanced R - choices section](https://adv-r.hadley.nz/control-flow.html#choices)

### Iteration

 * [R for Data Science - iteration chapter](https://r4ds.had.co.nz/iteration.html)
 * [Advanced R - loops section](https://adv-r.hadley.nz/control-flow.html#loops)


# Any questions?

