---
title: Introduction to conditional statements and loops in R
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
# These variables determine whether or not exercise solutions are included
show_solution <- TRUE # This determines if the solutions are displayed in the slides
purl_solutions <- FALSE # This variable relates to code blocks that are exercise solutions
purl_example_code <- TRUE # This variable relates to code blocks that aren't exercise solutions
```

## Setup

Before we start, make sure you've:

1. Got access to the `alpha-r-training` bucket
2. Cloned the `intro_r_training_extension` repo
3. Pulled any changes (the blue arrow on the Git tab in the top right panel)
4. Got `example_code.R` open
5. Opened the `intro_r_training_extension` repo on GitHub, so you can refer to the read me

If you need any help, just ask! For anyone joining remotely, we will be monitoring Teams throughout the session.


# Introduction

This Coffee and Coding session is trialing some new material that will build on the original [Introduction to R training course](https://github.com/moj-analytical-services/IntroRTraining), therefore we appreciate any feedback. It provides examples that demonstrate how the Tidyverse packages can assist with tasks typically encountered in MoJ Data & Analysis.

Development of the Tidyverse suite of packages was led by Hadley Wickham, and more information about these packages can be found on the [Tidyverse website](https://www.tidyverse.org/) as well as in the book [R for Data Science](https://r4ds.had.co.nz/).

## Learning outcomes

### By the end of this session you should know how to:

 - Classify a variable in a dataframe, based on a set of conditions
 - Easily apply a function to any number of columns in a dataframe


## Before we start

To follow along with the code run during this session and participate in the exercises, open the script "example_code.R" in RStudio. All the code that we'll show in this session is stored in "example_code.R", and you can edit this script to write solutions to the exercises. You may also want to have [this Markdown document](https://github.com/moj-analytical-services/intro_r_training_extension) with the session content open as a reference.

First, we need to load a few packages.

```{r message=F, warning=F, purl=purl_example_code}
# Load packages
library(botor)
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
```



```{r source_content, child = '../rmd_files/conditional_statements.Rmd'}
```

```{r source_content, child = '../rmd_files/iteration.Rmd'}
```


# Further Reading

## Further Reading

### General

 * [Tidyverse website](https://www.tidyverse.org)
 * [Tidyverse style guide](https://style.tidyverse.org/) (has some guidance on choosing function and argument names)
 * [MoJ coding standards](https://moj-analytical-services.github.io/our-coding-standards/)
 
### Conditional statements

 * [Advanced R - choices section](https://adv-r.hadley.nz/control-flow.html#choices)

### Iteration

 * [R for Data Science - iteration chapter](https://r4ds.had.co.nz/iteration.html)
 * [Advanced R - loops section](https://adv-r.hadley.nz/control-flow.html#loops)


# Any questions?

